---
title: "Extreme phenotype sampling"
author: "Thea Bjornland"
date: "20 June 18"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Extreme phenotype sampling}
  %\VignetteEncoding{UTF-8}
---

This package provides functions for testing hypothesis of genetic association with a continously measurable trait when data are from the extreme phenotype sampling (EPS) design. Functions for fitting a linear model are also provided.

```{r}
    library(extremesampling)
```

The assumed linear regression model is of the form
$$\mathbf{y} = X \beta + G \beta_g + \varepsilon,$$
where $y$ is a phenotype vector (continuosly measurable trait), $X$ is a matrix of 
non-genetic/environmental covariates (including intercept) and $G$ is a matrix 
of genetic covariates. The vector $\varepsilon$ is multivariate normal
with mean $0$ and covariance $I\sigma^2$ ($I$ is the identity matrix).


# Hypothesis tests

We have implemented score tests for testing $H_0: \beta_g = 0$ against $H_0: \beta_g \neq 0$, for two special situations

1. Test one genetic covariate (one column of $G$) at a time (the number of columns of $G$ can then be very large).
2. Test all genetic variants simultaneously (then then number of columns of $G$ should be relatively small).


We consider two types of data analysis

1. Complete case analysis. The response $y_i$, non-genetic covariates $\mathbf{x}_i$ and genetic covarirates $\mathbf{g}_i$ are observed only for extreme phenotype individuals, i.e. individuals who satisfy $y_i < l_i$ or $y_i > u_i$, where thresholds $l_i$ and $u_i$ are known, but might be different for each individual in the sample.

2. All case analysis. The response $y_i$ and non-genetic covariates $\mathbf{x}_i$ are observed for a full sample (randomly drawn from infinite population) while the genetic covarirates $\mathbf{g}_i$ are observed only for extreme phenotype individuals.


For complete case analysis, the thresholds must be supplied by the user. 

For all case analysis, the genotype matrix (or vector) must be of the same length as $\mathbf{y}$ and coded as NA where genotypes are not observed (non-extreme individuals). Further, it is assumed that the genotype is a discrete variable. The genotype distribution can be group-specific, and a matrix (dummy coded categorical covariates) that indicates these groups must be supplied. 

## Example data

We simulate data to illustrate score test functions. Extreme phenotype thresholds are set accoriding to the quantiles of the residual phenotype distribution ($z$-extreme sampling).


```{r}
    N = 5000 # Number of individuals in a population
    xe1 = rnorm(n = N, mean = 2, sd = 1) # Environmental covariate
    xe2 = rbinom(n = N, size = 1, prob = 0.3) # Environmental covariate
    Gmat = cbind(rbinom(N,2,0.4),rbinom(N,2,0.1),rbinom(N,2,0.25),
                 rbinom(N,2,0.01),rbinom(N,2,0.01),rbinom(N,2,0.01)) # Genetic variants
    
    # Generate phenotype
    betag = c(0.1,-0.1,0.3,0.5,-0.2,0)
    y = rnorm(N, mean = 10+5*xe1+1*xe2+Gmat%*%betag, sd = 4)
    
    # Identify extremes, here upper and lower 25% of residual phenotype distribution
    z = lm(y~xe1+xe2)$residuals
    uz = quantile(z,probs = 3/4,na.rm=TRUE)
    lz = quantile(z,probs = 1/4,na.rm=TRUE)
    extreme = (z < lz) | (z >= uz)
    li = lz + (y[extreme]-z[extreme]); ui = uz + (y[extreme]-z[extreme])
    
    # Create the EPS complete case data set
    yCC = y[extreme]; zCC = z[extreme]
    xe1CC = xe1[extreme]; xe2CC = xe2[extreme]
    GmatCC = Gmat[extreme,]
    
    # Create the EPS all case data set
    yAC = y; zAC = z
    xe1AC = xe1; xe2AC = xe2
    GmatAC = Gmat; GmatAC[!extreme,] = NA
```

## Single variant association tests

Each of the six columns of the genetic covariate matrix $G$ are tested one at a time 
against the null model $\mathbf{y} = X\beta + \varepsilon$. 



### Complete case tests

```{r,results='hide',message = FALSE}
cc = epsCC.test(yCC~xe1CC + xe2CC, xg = GmatCC, l = li, u = ui)
```
```{r}
print(cc)
```

### All case tests


```{r,results='hide',message = FALSE}
ac = epsAC.test(yAC~xe1AC + xe2AC, xg = GmatAC)
```

```{r}
print(ac)
```



## Multiple variant association tests

### Complete case tests

```{r,results='hide',message = FALSE}

# Direct test of all 6 variants
cc_d = epsCC.rv.test(yCC~xe1CC + xe2CC, xg = GmatCC, l = li, u = ui)$p.value

# Collapsing test (weights = 1 for all variants)
cc_c = epsCC.rv.test(yCC~xe1CC + xe2CC, xg = GmatCC, l = li, u = ui, method = "collapse")$p.value

# Variance component test (weights = 1 for all variants)
cc_v = epsCC.rv.test(yCC~xe1CC + xe2CC, xg = GmatCC, l = li, u = ui, method = "varcomp")$p.value

ccall = data.frame(cbind(cc_d, cc_c, cc_v))
names(ccall) = c("CC", "CC collapse", "CC varcomp")
```

```{r}
print(ccall)
```


### All case tests

```{r,results='hide',message = FALSE}

# Direct test of all 6 variants
ac_d = epsAC.rv.test(yAC~xe1AC + xe2AC, xg = GmatAC)$p.value

# Collapsing test (weights = 1 for all variants)
ac_c = epsAC.rv.test(yAC~xe1AC + xe2AC, xg = GmatAC, method = "collapse")$p.value

# Variance component test (weights = 1 for all variants)
ac_v = epsAC.rv.test(yAC~xe1AC + xe2AC, xg = GmatAC, method = "varcomp")$p.value

acall = data.frame(cbind(ac_d, ac_c, ac_v))
names(acall) = c("AC", "AC collapse", "AC varcomp")
```

```{r}
print(acall)
```


# Regression models

## Example data

We simulate data to illustrate score test functions. Extreme phenotype thresholds are equal for all individuals


```{r}
    N = 5000 # Number of individuals in a population
    xe1 = rnorm(n = N, mean = 2, sd = 1) # Environmental covariate
    xe2 = rbinom(n = N, size = 1, prob = 0.3) # Environmental covariate
    Gmat = cbind(rbinom(N,2,0.4),rbinom(N,2,0.1)) # Genetic variants
    
    # Generate phenotype
    betag = c(0.5,1)
    y = rnorm(N, mean = 10+5*xe1+1*xe2+Gmat%*%betag, sd = 4)
    
    # Identify extremes, here upper and lower 25% of residual phenotype distribution
    u = quantile(y,probs = 3/4,na.rm=TRUE)
    l = quantile(y,probs = 1/4,na.rm=TRUE)
    extreme = (y < l) | (y >= u)

    # Create the EPS complete case data set
    yCC = y[extreme]
    xe1CC = xe1[extreme]; xe2CC = xe2[extreme]
    GmatCC = Gmat[extreme,]
    
    # Create the EPS all case data set
    yAC = y
    xe1AC = xe1; xe2AC = xe2
    GmatAC = Gmat; GmatAC[!extreme,] = NA
```

### Complete case analysis

```{r,results='hide',message = FALSE}
cclm = epsCC.lm(yCC~xe1CC + xe2CC + GmatCC, l = li, u = ui)
```

```{r}
print(cclm$coefficients)
```

### All case analysis

```{r,results='hide',message = FALSE}
aclm = epsAC.lm(yAC~xe1AC + xe2AC + GmatAC)
```

```{r}
print(aclm$coefficients)
```

